Writing Trash Bin

### 2024-02-22 paper.qmd
updated task specific section

The idea that risk and ambiguity aversion reveal the utility of decreased variance seems to seldom be considered in the literature, therefore it should be mentioned here, otherwise this issue might continue. However, considering risk aversion is at its core, a trade off between bias and variance, it is bizzare to assume it to be a mistake when statisticians perform these trade offs all the time. If risk-aversion is truly "irrational" behaviour, every statistical textbook that discusses the use of a biased but consistent estimator in a finite sample case (a common example would be specific cases of the maximum likelihood estimator) should be tossed out. Discussing risk and ambiguity aversion as though it is some diversion from a perfectly rational, mathematical choice, implies there is *no value to a decrease in variance which is not even true within mathematics*. The concept of risk aversion treats risk as "noise", as though it is something that only exists to be ignored, but if the common place use of biased MLE's is an indication of anything, it is that *ignoring uncertainty* is the irrational choice.

@@

@@

## Perception Experiments, Question Problems, and Task Specificity
Uncertainty, like any other variable we are going to plot, has a natural mappings that should be considered when we express it in a plot. In the same way that mapping the colours red to hot and blue to cold, many variables, including uncertainty, have an intuitive mapping.

Even the most simple experiments are not free from the issues in the uncertainty definition.
Some experiments directly relate the internal 
Papers that consider unique aspects of uncertainty in their experiment such as experiemnts that identify the usefulness of specific perceptual tasks, experiments that identify problematic heuristics in commonly used uncertainty visualisations, and experiments that consider unique problems in uncertainty experiments all have useful results. These papers will be discussed in the upcomming sections.

Perceptual task experiments that identify visual aesthetics that are intuitively connected to uncertainty are useful as they establish an inutive psychological connection. These papers are also not without their flaws, however their flaws have little to do with issues in the grammar of graphics as they are well defined in this space and more to do with the lack. Since uncertainty and signal are typically mapped to different aesthetics, it is not always clear what exactly the visualisation is suggesting we should be uncertain about, and when a perceptual task that should be considered something like a train line has "uncertainty" added to it, it is not

>>>>

The idea that risk and ambiguity aversion reveal the utility of decreased variance seems to seldom be considered in the literature, therefore it should be mentioned here, otherwise this issue might continue. However, considering risk aversion is at its core, a trade off between bias and variance, it is bizzare to assume it to be a mistake when statisticians perform these trade offs all the time. If risk-aversion is truly "irrational" behaviour, every statistical textbook that discusses the use of a biased but consistent estimator in a finite sample case (a common example would be specific cases of the maximum likelihood estimator) should be tossed out. Discussing risk and ambiguity aversion as though it is some diversion from a perfectly rational, mathematical choice, implies there is *no value to a decrease in variance which is not even true within mathematics*. There have been other discussions on the appropriateness of discussing risk-aversion as a bias [@Vranas2000], but few have made the point that if uncertainty holds some *value* there is not technically a "right" decision at all. The concept of risk aversion treats risk as "noise", as though it is something that only exists to be ignored, but if the common place use of biased MLE's is an indication of anything, it is that *ignoring uncertainty* is the irrational choice.

@@

## Uncertainty and Task Specificity
When we consider mapping uncertainty, a reasonable questions that should come to mind is "uncertainty about what?". Every single thing concept or activity in statistics does not have uncertainty. For example, descriptive statistics cannot have uncertainty as they are just describing statistics as they are, the sample mean is the sample mean, we are not uncertain about its value. A sample variance describes dispersion, which is also how we describe uncertainty, but the sample variance in of itself does not have any uncertainty. Most of the mathematical uncertainty concepts have this idea embedded in them. For example, a confidence interval is an interval drawn around the sample mean that will theoretically contain the *populaton mean* 95% of the time. This issue has been noticed several times, however it is often brought up as a *task* dependence rather than an understanding that uncertainty is inherrently linked to inference on a particular statistic. @Wallsten1997 argue that the best method for evaluating or combining subjective probabilities depends on the uncertainty the decision maker wants to represent and why it matters. @Munzner2009 created a nested model for visualisation that highlighted how the first mistake that can be made in a visualisation is in the problem characterisation, and failling to do it well can cause downstream effects and damage the effectivness of a visualisation. @Fischhoff2014 looks at uncertainty visualisation for decision making decides that we should have different ways of communicating uncertainty based off what the user is supposed to do with it. This should make it clear that uncertainty only exists when we are performing inference. Therefore it only makes sense to describe uncertainty in terms of something to be uncertain *about*, it is not some a latent feature of statistics and data.

We feel the need to make this point clear because a *large* amount of uncertainty research seems to be confused by the purpose of uncertainty visualisation, what it is related to, and why we visualise it. Often, the authors are even more confused about what they are doing than the participants. There are enough papers that fail to understand these basic principals of uncertainty that this entire paper could simply be a list with explanations. Instead, we have a handful of highlights presented below.
- @Zhao2023 displayed a model prediction with uncertainty and took participants using the model prediction as a sign of trust, failing to notice that there is no reason for uncertainty information to influence someone to *use their own prediction*. If the Beuro of Meterology was uncertain about their rain prediction, that would not be sufficient conditions for me to decide to make my own rain predictions. Despite this, the authors seemed to assume that the uncertainty information *should* do that.  
- @Hofman2020 asked questions about discusses the difference in showing inferential or outcome (prediction) intervals and then found that participants shown the inferential distribution performed poorly when asked questions about predictions. They found that showing the relevant distribution to the question reduced the error. This result should be astoundingly obvious if it was well understood that uncertainty is not some latent variable and is only relevant for specific ground truth statistics. This papers *percieved* importants highlights an underlying belief that uncertainty is linked to a specific variable or data set, not the variable we are drawing inference on.
- @Padilla2021 found that high uncertainty in the model estimates (calculated uncertainty) and low forecaster confidence (which is typically an expression of suspected bias in the model) both caused participants to have decreased confidence in their results and suggested modelers express both if they are relevant. @Kale2019 discussed the importance of communicating decisions made in the data analysis pipeline and being aware of the alternatives. While there is nothing wrong with the core ideas of these papers, the fact that they were published indicates that the general population was not already aware that choices introduced early in the data analysis pipeline create bias and therefore uncertainty in our final values. This indicates a general failure to see bias and variance both as sources of uncertainty.
- @Wickham2011 suggests their product plot framework, which includes histograms, should have a way to measure uncertainty, but does not consider that a histogram is *already* a depiction of mass and would already be considered an uncertainty visualisation were our statistic of interest the population mean. To suggest that a histogram needs uncertainty bounds is similar to suggesting a density or box plot need uncertainty bounds. 
- @Griethe2006 commented that "if visualization is used as a means to explore a data volume or to communicate its contents the uncertainty has to be included". @Potter2010 aimed to create a summary plot that "concisely presented data with uncertaninty information". These papers fail to realise the exploration step of an analysis, which includes descriptive statistics, exploratory data visualisaiton, and unsupervised machine learning techniques, is performed without a prior hypothesis. Therefore, it is performed to *suggest* a direction for future inference work or communication, which *do* contain uncertainty and does not contain uncertainty within itself. This confusion would not occur if it was understood that uncertainty does not exist without inference.
- @Ibrekk1987 asked participants for the "best" estimate while displaying a population density function. The participants proided the mode instead of the mean, although there was no indication the mean was what was being asked about.
- @Boukhelifa2012 tried to quantify the strength of the intuitive connection between "sketchiness" and uncertainty. Since they had seen literature that explained uncertainty was task specific, they tested the connection for multiple "tasks", what they failled to understand, is that the "task" dependency which is likely just a proxy of the evaluation experiment problem discussed in the previous section. This misunderstanding led to "sketchiness" being used to depict uncertainty in the categories of bar charts, in graphic networks, and in train lines, with little consideration or explanation as to *how* these things would be uncertainty. I imagine it is quite hard to be "uncertain" about the existence of a train line. Participants rightfully assumed the sketchiness therefore represented something else, such alternative options or simply ignored it.

![The graphics displayed by @Boukhelifa2012 to identify if there is an intuitive connection between sketchiness and uncertainty.](sketchiness.png){#fig-sketchy}

The reality is that a lot of uncertainty visualisation authors do not seem to have an intuitive understanding of uncertainties connection to inference, when inference is being performed, and how to design experiments that capture this relationship.

This misunderstanding is not due to laziness or the fault of the authors, but rather is likely caused by the absence of a strict definition of uncertainty. The imprecise and confusing set of existing definitions are creating a field in which the authors themselves do not know what they are testing. They do not even seem to understand what uncertainty is. Uncertainty examples include imputed data, model selection, inherrent randomness, biased sampling, etc, not because these things *are* uncertainty, but because they *create* uncertainty when we perform inference. These things are not "uncertainty" in of themselves, but rather contribute to the distance between the final estimate and the statistic we want to perform inference on. The field of uncertainty communication is in desperate need of a unifying mathematical definition that will allow the bias and variance introduced through various stages of the data analysis pipeline to be compared or quantified.

Most uncertainty definitions speak about this vague concept of uncertainty but do not

The fact that uncertainty only exists when we are performing inference also creates issues with the questions of an experiment. If uncertainty is depicted and uncertainty is always related to . The authors of these experiments often comment on the results from the perspective that the participants have made a conceptual mistake, and at no point do they question why we would depict uncertainty for an unrelated ground truth. The results of these experiments essentially show that participants can be "tricked" by failling to understand uncertainty (showing that people can be tricked if you purposely lead them astray should not constitute interesting research). Ironically the authors fail to realise they have made this same mistake when they did not consider displaying uncertainty for a fundamentally unrelated

assume the participants have made some sort of mistake, rather than questioning their own understanding of uncertainty.

must be related to the statistic we wish to draw inference on

Reguardless of whether or not the visualisation and associated uncertainty is related to the statistic we are trying to draw inference from, participants will *assume* they are.

@Hofman2020 suggests authors depictions of the

This concept is the first issue that uncertainty experiments run into, specifically in perceptual task experiments.

It is rarely mentioned in the uncertainty literature, but

Even the most simple experiments are not free from the issues in the uncertainty definition.
Some experiments directly relate the internal

@@

### 2024-02-22 writingdumpster.qmd
updated task specific section

He kept the taxonomy to three levels to maintain simplicity, however this organisation methods conflates amount of uncertainty present in an estimate with what is causing that uncertainty, and in doing so makes it hard for us to use this definition to better understand uncertainty within our own analysis.
# context section outtakes
Given that uncertainty estimation and communication seem to be largely task depenent, what questions should we be testing? This leads to several unanswered questions. Is our goal in uncertainty communication to acknowledge and convey the inherrent randomness that exists in all statistical calculations and models? Is it to supress signal that could be the result of random variance? Is it to find the best way to convey the mass function of a random variable? A survey of the literature will leave the answers to these questions unclear, however this inherrent issue has not gone unnoticed.
This conceptual way of thinking about uncertainty has been rattling around in the different areas of data science for a while.
For example, when we record coin flips we typically want to model the behavior with a binomial distribution, so we ignore the outcome of a coin landing on its side despite the fact it is a real world possibility.
@Wallsten1997 argue that the best method for evaluating or combining subjective probabilities depends on the uncertainty the decision maker wants to represent and why it matters.  @Munzner2009 created a nested model for visualisation that highlighted how pitfalls in upstream levels, such as problem characterisation and data abstraction, can cause downstream effects and damage the effectivness of a visualisation. @Fischhoff2014 looks at uncertainty visualisation for decision making decides that we should have different ways of communicating uncertainty based off what the user is supposed to do with it.

@@

# Previous literature review
#####Taxonomy Benefits
- Distinctions at this taxonomy level seem to be important also untangling helps to understand it
  - @Padilla2021 found that low forecaster confidence or high model uncertainty both contribute to more conservative judgements by decision makers.
  - Failing to communicate the nature of your uncertainty can result in underestimation or overestimation of failure probabilities [@Kiureghian2009]
  - @Gustafson2019 found that the framing of our uncertainty, (i.e. if the source of uncertainty is from a lack of knowledge, approximations, unknown unknowns, or disagreement among parties) was found to not have a detrimental effect on the belief in the estimates, perceived credibility, or behavioural intentions of the decision makers.

>>>>

He kept the taxonomy to three levels to maintain simplicity, however this organisation methods conflates amount of uncertainty present in an estimate with what is causing that uncertainty, and in doing so makes it hard for us to use this definition to better understand uncertainty within our own analysis.
# context section outtakes
Given that uncertainty estimation and communication seem to be largely task depenent, what questions should we be testing? This leads to several unanswered questions. Is our goal in uncertainty communication to acknowledge and convey the inherrent randomness that exists in all statistical calculations and models? Is it to supress signal that could be the result of random variance? Is it to find the best way to convey the mass function of a random variable? A survey of the literature will leave the answers to these questions unclear, however this inherrent issue has not gone unnoticed.
This conceptual way of thinking about uncertainty has been rattling around in the different areas of data science for a while.
For example, when we record coin flips we typically want to model the behavior with a binomial distribution, so we ignore the outcome of a coin landing on its side despite the fact it is a real world possibility.

@@

# Previous literature review
#####Taxonomy Benefits
- Distinctions at this taxonomy level seem to be important also untangling helps to understand it
  - Failing to communicate the nature of your uncertainty can result in underestimation or overestimation of failure probabilities [@Kiureghian2009]
  - @Gustafson2019 found that the framing of our uncertainty, (i.e. if the source of uncertainty is from a lack of knowledge, approximations, unknown unknowns, or disagreement among parties) was found to not have a detrimental effect on the belief in the estimates, perceived credibility, or behavioural intentions of the decision makers.
